Dark matter halo shape in the Auriga simulations: radial dependence, time evolution and correlation with baryon disk properties

### 2019-07-14 Document/paper.tex
edit method

The Auriga project offers cosmological zoom in simulations of MW-sized 
dark matter halos in a $\Lambda CDM$ cosmology. 
This simulations come in two versions: dark matter only and
baryonic physics including magetohydrodynamics (MHD).
A detailed description of the simulations and their
disk properties can be founc in \citep{auriga}, here we summarize its
main features.

@@

The 30 zoom-in halos have a dark matter particle mass of $\sim 3\times
10^5$\Msun while the barynic mass resolution is $\sim 5\times 10^4$\Msun.
The softening lenght for gravitational force computation for stellar
particles and high-resolution dark matter particles 
is fixed to be 500 $h^{-1}$ pc in comoving coordinates up to $z=1$,
and 396 pc in physical coordinates afterwards.
The gravitational softening lenght for gas cells changes with the mean
cell radius but is limited to be larger than the stellar softening
lenght and 1850 pc physical. 
This setup corresponds to \texttt{Level4} resolution.
From these 30 halos, 6 of them where re-simulated at higher resolution
taking into account a spatial factor of 2 in each spatial dimension,
this is the \texttt{Level3} resolution.

In this work we analyse all the 30 halos in the \texttt{Level4}
resolution. 
All the results that we report at $z=0$ as a function of radius and
the correlations with disk properties are performed with that sample.
For the measurement on shape evolution with time we use the 6 halos in
the \texttt{Level3} simulations.
Finally, all  halos described so far have also been simulated at the same
resolutions without MHD using dark matter particles only, we refer to
these halos as the DMO (Dark Matter Only) sample.

@@

The DM halo shape at a fixed radius is an estimate of either
the isopotential or isodensity surfaces.  
Observational inference models usually estimate the 
isopotential contours which are probed by tracers (gas, stars), while
simulations work with the isodensity contours which can be directly
calculated from particle positions.

However, in numerical simulations the density contours are not smooth
and are sensitive to the presence of small satelites. 
For this reason we choose to measure the shape by taking
volume-enclosed particles, rather than shell-enclosed.  
We follow the shape measurement method presented by
\cite{Allgood_et_al._2006} that uses the reduced
inertia tensor,

@@

with the positions components weighted by the k-th particle distance
$d_k^2=x_k^2+y_k^2+z_k^2$, the particle positions are measured from
the minimum of the gravitational potential in each halo.
The diagonalization of this tensor yields the principal axes of the
structure as well as the eigen-quantities which are proportional to
the squared principal axes $a>b>c$.

@@

We restrict the sampling of the ellipsoidal parameters to radii
between $1/16 R_{vir}$ and $2R_{vir}$, where  $R_{vir}$ is taken as the
radius enclosing a sphere with 200 times the average dark matter
density of the Universe. 
All our results use of this reference radius
unless strictly stated otherwise. 
We peform the shape measurements both as a function of radius and
redshift for all halos in the sample.

>>>>

The Auriga project offers cosmological zoom in simulations of MW-sized 
dark matter halos in a $\Lambda$CDM cosmology. 
This simulations come in two versions: dark matter only and
baryonic physics including magetohydrodynamics (MHD).
A detailed description of the simulations and their
disk properties can be founc in \citep{auriga}, here we summarize its
main features.

@@

The 30 zoom-in halos have a dark matter particle mass of $\sim 3\times
10^5$\Msun while the barynic mass resolution is $\sim 5\times 10^4$\Msun.
The softening lenght for gravitational force computation for stellar
particles and high-resolution dark matter particles 
is fixed to be 500 $h^{-1}$ pc in comoving coordinates up to $z=1$,
and 396 pc in physical coordinates afterwards.
The gravitational softening lenght for gas cells changes with the mean
cell radius but is limited to be larger than the stellar softening
lenght and 1850 pc physical. 
This setup corresponds to \texttt{Level4} resolution.
From these 30 halos, 6 of them where re-simulated at higher resolution
taking into account a spatial factor of 2 in each spatial dimension,
this is the \texttt{Level3} resolution.  
There are $\sim 4\times 10^6$ dark matter particles within the virial radius
of \texttt{Level4} halos at $z=0$, this number increases to $\sim 3\times 10^7$ in
\texttt{Level3} simulations.

In this work all the results that we report at $z=0$ as a function of radius and
the correlations with disk properties correspond to the 30 halos in
the \texttt{Level4} resolution. 
For the measurement on shape evolution with time we use the 6 halos in
the \texttt{Level3} simulations.
Finally, all  halos described so far have also been simulated at the same
resolutions without MHD using dark matter particles only, we refer to
these halos as the DMO (Dark Matter Only) sample.

@@

The DM halo shape at a fixed radius is an estimate of either
the isopotential or isodensity surfaces.  
Observational inference models usually estimate the 
isopotential contours which are probed by tracers (gas, stars), while
simulations work with the isodensity contours which can be directly
calculated from particle positions.  
Furthermore, the density contours in thin shells are very sensitive to
the presence of small satelites.  
For this reason we measure the shape by taking
volume-enclosed particles, rather than shell-enclosed.

We measure the shape using the reduced inertia tensor \citep{Allgood_et_al._2006},

@@

where the particle positions are measured from the minimum of the
gravitational potential in each halo and each is weighted by the k-th
particle distance 
$d_k^2=x_k^2+y_k^2+z_k^2$,

The diagonalization of this tensor yields the eigenvectors and
eigenvalues that represent an ellipsoidal dark matter halo.
The axis lenghts of this ellipsoid $a\geq b \geq c$ are the square
root of the ${\bf I}$ eigenvalues and the direction of the principal
axis are the corresponding eigenvectors

@@

Following the convergence critery by \cite{Vera-Crito_et_al.2011} we
restrict the sampling of the ellipsoidal parameters to radii  between
$\sim 2$kpc and $R_{200}$, where  $R_{200}$ correspond to the  radius enclosing a sphere with 200 times the average dark matter
density of the Universe. 
On average, over the 30 halos in the \texttt{Level4} sample
$R_{200}=200\pm 10$kpc. 
For \texttt{Level3} halos we go down to distances $\sim 0.2$ kpc.
